---
title: "Customer Personality Analysis"
subtitle: "Segmenting and predicting customer behavior"
author: "Ciaboco Alessandro &nbsp;·&nbsp; Matricola 68530A"
date: "May 2026"
output:
  ioslides_presentation:
    widescreen: true
    self_contained: true
---

<style>
@import url('https://fonts.googleapis.com/css2?family=Inter:wght@300;400;500;600;700&display=swap');
body, slides > slide { font-family: 'Inter', 'Segoe UI', sans-serif; }
slides > slide { background: #F8FAFC; color: #1C2B4A; }

slides > slide.title-slide { background: #1C2B4A; border-left: 6px solid #0D9488; }
slides > slide.title-slide hgroup h1 { color:#FFF; font-size:40px; font-weight:700; }
slides > slide.title-slide hgroup h2 { color:#5EEAD4; font-size:18px; font-weight:400; border:none; }
slides > slide.title-slide p { color:#94A3B8; font-size:14px; }

slides > slide.segue { background:#1C2B4A !important; border-left:8px solid #0D9488; }
slides > slide.segue hgroup h2 { color:#FFF !important; font-size:36px !important; }
slides > slide.segue hgroup h3 { color:#5EEAD4 !important; }

slides > slide:not(.title-slide):not(.segue) > hgroup h2 {
  font-size:26px; font-weight:700; color:#1C2B4A;
  border-bottom:3px solid #0D9488; padding-bottom:6px; margin-bottom:16px;
}

.teal { color:#0D9488; font-weight:600; }
.navy { color:#1C2B4A; font-weight:600; }
.muted { color:#64748B; font-size:.9em; }
.center { text-align:center; }

.card {
  background:#FFF; border:1px solid #E2E8F0; border-radius:8px;
  padding:12px 16px; margin-bottom:10px; box-shadow:0 1px 4px rgba(0,0,0,.06);
}
.card-teal  { border-left:4px solid #0D9488; }
.card-navy  { border-left:4px solid #1C2B4A; }
.card-amber { border-left:4px solid #B45309; }
.card-purple{ border-left:4px solid #7C3AED; }
.card h4 { margin:0 0 5px; font-size:13px; font-weight:600; color:#1C2B4A; }
.card p  { margin:0; font-size:12px; color:#64748B; line-height:1.5; }

.col2 { display:grid; grid-template-columns:1fr 1fr; gap:18px; align-items: start; }
.col3 { display:grid; grid-template-columns:1fr 1fr 1fr; gap:12px; }
.grid-60-40 { display:grid; grid-template-columns: 1.3fr 0.7fr; gap: 20px; align-items: center; }

.hi {
  background:#CCFBF1; border-left:4px solid #0D9488; border-radius:6px;
  padding:10px 14px; margin:8px 0; font-size:12px; color:#0F766E; line-height:1.5;
}
.hi-warn { background:#FEF3C7; border-left:4px solid #B45309; color:#92400E; }
.hi-red  { background:#FEE2E2; border-left:4px solid #DC2626; color:#991B1B; }

.stat-row { display:flex; gap:12px; margin-bottom:12px; }
.stat-box {
  flex:1; background:#1C2B4A; border-radius:8px;
  padding:14px 10px; text-align:center; color:#FFF;
}
.stat-box .num { font-size:28px; font-weight:700; color:#0D9488; line-height:1; display:block; }
.stat-box .lbl { font-size:10px; color:#94A3B8; margin-top:3px; display:block; }

.pill { display:inline-block; padding:3px 10px; border-radius:99px; font-size:11px; font-weight:600; margin:2px; }
.pill-teal  { background:#CCFBF1; color:#0F766E; }
.pill-red   { background:#FEE2E2; color:#991B1B; }
.pill-navy  { background:#DBEAFE; color:#1E40AF; }
.pill-amber { background:#FEF3C7; color:#92400E; }

.cm-wrap { display:flex; justify-content:center; margin:6px 0; }
.cm-table { border-collapse:collapse; font-size:13px; width:100%; }
.cm-table th { background:#1C2B4A; color:#FFF; padding:8px 18px; text-align:center; }
.cm-table td { padding:10px 18px; text-align:center; border:1px solid #E2E8F0; font-weight:600; }
.cm-tp { color:#166534; background:#DCFCE7; }
.cm-tn { color:#166534; background:#DCFCE7; }
.cm-fn { color:#991B1B; background:#FEE2E2; }
.cm-fp { color:#92400E; background:#FEF3C7; }

.ag-item {
  display:flex; align-items:center; padding:9px 14px; margin-bottom:7px;
  border-radius:8px; background:#FFF; border:1px solid #E2E8F0;
  font-size:14px; color:#1C2B4A;
}
.ag-num { font-size:22px; font-weight:700; color:#0D9488; min-width:40px; }

.step {
  display:inline-flex; align-items:center; justify-content:center;
  width:24px; height:24px; border-radius:50%;
  background:#0D9488; color:#FFF; font-size:12px; font-weight:700;
  margin-right:8px; flex-shrink:0;
}

.pipe { display:flex; align-items:stretch; gap:6px; margin-top:12px; }
.pipe-box {
  flex:1; background:#FFF; border-radius:8px;
  border:1px solid #E2E8F0; padding:12px 10px; text-align:center;
}
.pipe-box h4 { color:#1C2B4A; font-size:12px; font-weight:600; margin:0 0 4px; }
.pipe-box p  { color:#64748B; font-size:11px; margin:0; line-height:1.4; }
.pipe-arrow  { display:flex; align-items:center; color:#0D9488; font-size:20px; font-weight:700; }

.cl-table { width:100%; border-collapse:collapse; font-size:12px; margin-top:8px; }
.cl-table th { background:#1C2B4A; color:#FFF; padding:7px 10px; text-align:left; }
.cl-table td { padding:7px 10px; border-bottom:1px solid #E2E8F0; color:#334155; }
.cl-table tr:nth-child(even) td { background:#F1F5F9; }

slides > slide:not(.title-slide)::after {
  content:'Ciaboco Alessandro · Customer Personality Analysis · May 2026';
  position:absolute; bottom:8px; right:16px; font-size:9px; color:#CBD5E1;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, fig.align="center")

library(dplyr)
library(ggplot2)
library(factoextra)
library(caret)
library(glmnet)
library(randomForest)
library(nnet)
library(knitr)
library(tidyr)

df <- read.csv("marketing_campaign.csv", sep="\t", stringsAsFactors=TRUE)

df <- df %>%
  mutate(
    Age          = 2026 - Year_Birth,
    TotalSpent   = MntWines + MntFruits + MntMeatProducts +
                   MntFishProducts + MntSweetProducts + MntGoldProds,
    ResponseTarget = as.factor(ifelse(Response == 1, "Yes", "No"))
  ) %>%
  filter(Income < 600000, Age < 100) %>%
  drop_na(Income)

df_clean <- df %>%
  select(-ID, -Z_CostContact, -Z_Revenue, -Year_Birth, -Dt_Customer, -Response)

num_vars    <- df_clean %>% select(where(is.numeric)) %>%
               select(-starts_with("Accepted"), -Complain)
scaled_data <- scale(num_vars)
pca_res     <- prcomp(scaled_data, center=FALSE, scale.=FALSE)

set.seed(123)
km_res   <- kmeans(scaled_data, centers=3, nstart=25)
dist_mat <- dist(scaled_data, method="euclidean")
hc_res   <- hclust(dist_mat, method="ward.D2")

set.seed(123)
trainIndex <- createDataPartition(df_clean$ResponseTarget, p=0.8, list=FALSE)
train_data <- df_clean[trainIndex,]
test_data  <- df_clean[-trainIndex,]
cv_ctrl    <- trainControl(method="cv", number=5,
                           classProbs=TRUE, summaryFunction=twoClassSummary)

set.seed(123)
log_model <- train(ResponseTarget~., data=train_data, method="glm",
                   family="binomial", metric="ROC", trControl=cv_ctrl)
set.seed(123)
lasso_model <- train(ResponseTarget~., data=train_data, method="glmnet",
                     tuneGrid=expand.grid(alpha=1, lambda=seq(0.01,0.1,0.02)),
                     metric="ROC", trControl=cv_ctrl)
set.seed(123)
rf_model <- train(ResponseTarget~., data=train_data, method="rf",
                  metric="ROC", trControl=cv_ctrl, tuneLength=2)
set.seed(123)
ann_model <- train(ResponseTarget~., data=train_data, method="nnet",
                   metric="ROC", trControl=cv_ctrl, trace=FALSE, tuneLength=2)

pred_rf  <- predict(rf_model, newdata=test_data)
conf_mat <- confusionMatrix(pred_rf, test_data$ResponseTarget, positive="Yes")

prob_rf  <- predict(rf_model, newdata=test_data, type="prob")[,"Yes"]
pred_adj <- factor(ifelse(prob_rf>=0.3,"Yes","No"), levels=c("No","Yes"))
conf_adj <- confusionMatrix(pred_adj, test_data$ResponseTarget, positive="Yes")

cl_prof <- df_clean %>%
  mutate(Cluster = km_res$cluster) %>%
  group_by(Cluster) %>%
  summarise(N=n(), Income=round(mean(Income),0), TotalSpent=round(mean(TotalSpent),0),
            Recency=round(mean(Recency),1), WebVisits=round(mean(NumWebVisitsMonth),1),
            .groups="drop") %>%
  arrange(desc(Income)) %>%
  mutate(Segment=c("Premium","Middle","Browsers")) %>%
  select(Segment, N, Income, TotalSpent, Recency, WebVisits)
```

## Agenda

<div class="ag-item"><span class="ag-num">01</span>&nbsp; Dataset &amp; Data cleaning</div>
<div class="ag-item"><span class="ag-num">02</span>&nbsp; Unsupervised: PCA &amp; Clustering</div>
<div class="ag-item"><span class="ag-num">03</span>&nbsp; Supervised: 4 models, comparison, test set</div>
<div class="ag-item"><span class="ag-num">04</span>&nbsp; Threshold tuning &amp; business implications</div>
<div class="ag-item"><span class="ag-num">05</span>&nbsp; Key takeaways</div>

## The problem &amp; the goal

<div class="col2">
<div>
<div class="card card-navy">
<h4>Business problem</h4>
<p>A retail company sends the same offer to all customers without any targeting. This wastes budget on people who won't convert and misses those most likely to buy.</p>
</div>
<div class="card card-teal" style="margin-top:10px">
<h4>Two-part approach</h4>
<p><b>Part 1:</b> PCA + K-Means to find natural customer segments.<br><br>
<b>Part 2:</b> Four classification models to predict who will accept the next campaign.</p>
</div>
</div>
<div>
<div class="stat-row">
<div class="stat-box"><span class="num">2,240</span><span class="lbl">customers</span></div>
<div class="stat-box"><span class="num">29</span><span class="lbl">variables</span></div>
</div>
<div class="stat-row">
<div class="stat-box"><span class="num">~15%</span><span class="lbl">positive response rate</span></div>
<div class="stat-box"><span class="num">5</span><span class="lbl">past campaigns</span></div>
</div>
<div class="hi">Cross-sectional dataset. One row per customer, iid. Mixed types: numeric, categorical, binary.</div>
</div>
</div>

## Dataset: Variable groups & features

<div class="col2">
<div>
<div class="card card-navy">
<h4>Demographics & Household</h4>
<p>Core customer profiling attributes used to establish baseline socioeconomic status.</p>
<div style="margin-top:8px;">
  <span class="pill pill-navy">Age</span>
  <span class="pill pill-navy">Income</span>
  <span class="pill pill-navy">Education</span>
  <span class="pill pill-navy">Marital_Status</span>
  <span class="pill pill-navy">Kidhome</span>
  <span class="pill pill-navy">Teenhome</span>
</div>
</div>
<div class="card card-teal">
<h4>Product Spending Profile</h4>
<p>Two-year historical spend data across six primary retail categories. Aggregated into a master <b>TotalSpent</b> feature to represent absolute buying power.</p>
<div style="margin-top:8px;">
  <span class="pill pill-teal">Wines</span>
  <span class="pill pill-teal">Fruits</span>
  <span class="pill pill-teal">Meat</span>
  <span class="pill pill-teal">Fish</span>
  <span class="pill pill-teal">Sweets</span>
  <span class="pill pill-teal">Gold</span>
</div>
</div>
</div>
<div>
<div class="card card-purple">
<h4>Channel Engagement & Traffic</h4>
<p>Interaction metrics detailing exactly how the customer shops. Separating visits from purchases is critical to identify browsers.</p>
<ul style="margin:8px 0 0 16px; padding:0; font-size:12px; color:#64748B;">
  <li><b>NumWebVisitsMonth:</b> Digital traffic proxy</li>
  <li><b>NumWebPurchases:</b> E-commerce conversion</li>
  <li><b>NumStorePurchases:</b> Physical retail</li>
  <li><b>NumCatalogPurchases:</b> Direct mail</li>
</ul>
</div>
<div class="card card-amber">
<h4>Campaign History & The Target</h4>
<p>Historical record of promotional acceptance across 5 previous marketing initiatives.</p>
<div style="margin-top:8px; padding-top:8px; border-top:1px solid #FDE68A;">
  <p style="color:#92400E;"><b>Target Feature:</b> <code>ResponseTarget</code> (Yes/No). Encodes the acceptance of campaign 6 and acts as the label for all supervised models.</p>
</div>
</div>
</div>
</div>

<div class="hi" style="margin-top:16px;">
<b>Feature Engineering Note:</b> Raw dates (e.g., <code>Year_Birth</code>) were transformed into continuous numerical features (<code>Age</code>) to ensure compatibility with distance-based algorithms like K-Means.
</div>

## Data Cleaning: Ensuring Result Integrity

<div class="col2">
<div>
<div class="card card-teal">
<h4><span class="step">1</span> Handling Missing Values</h4>
<p>Rows without income data were removed to prevent estimation bias. Imputation was avoided as it could introduce artificial signals in financial modeling.</p>
<div style="margin-top:8px;">
  <span class="pill pill-red">Removed: Income (NA)</span>
</div>
</div>
<div class="card card-teal">
<h4><span class="step">2</span> Outlier Suppression</h4>
<p>Strict filtering to stabilize variance. We capped Age at 100 and Income at 600k to remove obvious data entry errors.</p>
<div style="margin-top:8px;">
  <span class="pill pill-amber">Filtered: Age > 100</span>
  <span class="pill pill-amber">Filtered: Income > 600k</span>
</div>
</div>
</div>
<div>
<div class="card card-navy">
<h4><span class="step">3</span> Zero-Variance Features</h4>
<p>Variables with constant values across all observations provide zero predictive power. Their removal reduces computational noise.</p>
<div style="margin-top:8px;">
  <span class="pill pill-navy">Dropped: Z_CostContact</span>
  <span class="pill pill-navy">Dropped: Z_Revenue</span>
</div>
</div>
<div class="card card-navy">
<h4><span class="step">4</span> Administrative Noise</h4>
<p>Variables used for row identification or join dates were excluded to prevent models from learning coincidental patterns.</p>
<div style="margin-top:8px;">
  <span class="pill pill-navy">Dropped: ID</span>
  <span class="pill pill-navy">Dropped: Dt_Customer</span>
</div>
</div>
</div>
</div>

<div class="hi" style="margin-top:16px;">
<b>Methodological Rationale:</b> The cleaning phase focused on <b>Signal-to-Noise Ratio</b> improvement. By prioritizing data quality over quantity, we ensure that the subsequent PCA and predictive models are grounded in realistic behavioral patterns.
</div>

# Part 1: Unsupervised Learning {.segue}

## PCA: Visual Interpretation & Biplot

<div class="grid-60-40">
<div>
```{r pca-biplot, fig.height=5.2, fig.width=5.5}
fviz_pca_var(pca_res, col.var="contrib",
             gradient.cols=c("#00AFBB","#E7B800","#FC4E07"),
             repel=TRUE, select.var=list(contrib=10)) +
  labs(title=NULL) +
  theme_minimal(base_size=12) +
  theme(plot.background = element_blank(),
        panel.grid.major = element_line(color = "#E2E8F0"))
```
</div>
<div>
<div class="card card-navy">
<h4>PC1: Economic Power (42.1%)</h4>
<p>Income, TotalSpent, and MntWines all load in the same direction. A higher PC1 score equals a higher-value customer.</p>
</div>
<div class="card card-teal">
<h4>Web Visits vs Spending</h4>
<p><code>NumWebVisitsMonth</code> points directly opposite to spending features. Most frequent visitors tend to be the lowest spenders.</p>
</div>
<div class="card card-purple">
<h4>PC2: Family Structure (11.6%)</h4>
<p><code>Teenhome</code> loads heavily on PC2, capturing a household dimension independent of raw income.</p>
</div>
<div class="hi">
Segmentation relies heavily on spending levels, visually confirmed by the dominant PC1 axis.
</div>
</div>
</div>

## Clustering: Choosing optimal k

<div class="grid-60-40">
<div>
```{r elbow, fig.height=5, fig.width=5.5}
fviz_nbclust(scaled_data, kmeans, method="wss") +
  labs(title=NULL) +
  theme_minimal(base_size=14) +
  theme(plot.background = element_blank())
```
</div>
<div>
<div class="card card-teal">
<h4>The Elbow Method</h4>
<p>Evaluating the Within-Cluster Sum of Squares reveals a clear bend at <b>k = 3</b>.</p>
</div>
<div class="card card-navy">
<h4>Why stop at 3?</h4>
<p>Increasing the number of clusters beyond 3 flattens the curve, meaning we would add complexity without capturing fundamentally different consumer behaviors.</p>
</div>
</div>
</div>

## Hierarchical Dendrogram: Structure Validation

<div class="grid-60-40">
<div>
```{r dendrogram, fig.height=5.5, fig.width=4.5}
plot(hc_res, labels=FALSE, hang=-1,
     main="Ward.D2 Linkage", xlab="", sub="", cex.main=1.4)
rect.hclust(hc_res, k=3, border=c("#1C2B4A","#0D9488","#7C3AED"))
```
</div>
<div>
<div class="card card-purple">
<h4>Structural Confirmation</h4>
<p>The hierarchical approach validates our K-Means setup. The dendrogram naturally splits into three distinct main branches.</p>
</div>
<div class="card card-navy">
<h4>High Stability</h4>
<p>The large vertical gap between the 3-cluster split and the subsequent lower splits confirms that the chosen partition is robust and deeply rooted in the data.</p>
</div>
</div>
</div>

## Cluster Profiles: The Three Customer Tiers

```{r cluster-table, results='asis'}
kable(cl_prof,
      col.names=c("Segment","N","Avg Income (€)","Avg Total Spent (€)",
                  "Avg Recency (days)","Avg Web Visits/mo"),
      align="lrrrrr", format="html", table.attr='class="cl-table"')
```

<div class="col3" style="margin-top:20px;">
<div class="card card-purple">
<h4>Premium Segment</h4>
<p>Highest income and maximum total spend. They engage less with the website but convert at high values.</p>
</div>
<div class="card card-navy">
<h4>Middle Segment</h4>
<p>The mass market. Sits directly between the two extremes on all financial and engagement metrics.</p>
</div>
<div class="card card-teal">
<h4>Browser Segment</h4>
<p>Lowest absolute spend, but highest web visits. Represents a pool of latent value requiring conversion strategies.</p>
</div>
</div>

# Part 2: Supervised Learning {.segue}

## Setup

<div class="col2">
<div>
**Target Variable:** `ResponseTarget` (binary Yes/No), indicating campaign 6 acceptance.

<div class="hi-warn" style="margin-top:10px">
<b>Class imbalance:</b> ~15% positive responses. AUC-ROC was strictly chosen as the evaluation metric to prevent misleading accuracy scores.
</div>

<div class="card card-teal" style="margin-top:15px">
<h4>5-fold cross-validation</h4>
<p>Each fold is used once as validation. Prevents overfitting and provides stable AUC estimates prior to touching the holdout set.</p>
</div>
</div>
<div>
<div class="stat-row">
<div class="stat-box"><span class="num">80%</span><span class="lbl">training set</span></div>
<div class="stat-box"><span class="num">20%</span><span class="lbl">test set</span></div>
</div>
<div class="card card-navy" style="margin-top:8px">
<h4>Models evaluated</h4>
<p>Logistic regression</p>
<p>Lasso</p>
<p>Random Forest</p>
<p>Artificial Neural Network (ANN)</p>
</div>
</div>
</div>

## Feature Importance: Drivers of Conversion

<div class="grid-60-40">
<div>
```{r varimp, fig.height=5.2, fig.width=5.5}
vi <- varImp(rf_model)$importance %>%
  tibble::rownames_to_column("Feature") %>%
  arrange(Overall) %>% tail(10) %>%
  mutate(Feature=factor(Feature, levels=Feature))

ggplot(vi, aes(x=Feature, y=Overall)) +
  geom_segment(aes(xend=Feature, yend=0), color="#CBD5E1", linewidth=1.2) +
  geom_point(size=4.5, color="#0D9488") +
  coord_flip() +
  labs(title=NULL, x=NULL, y="Importance") +
  theme_minimal(base_size=14) +
  theme(plot.background = element_blank(), panel.grid.major.y=element_blank(), panel.grid.minor=element_blank())
```
</div>
<div>
<div class="card card-teal">
<h4>Recency is #1</h4>
<p>Days since last purchase dominates. Recent buyers are drastically more likely to convert.</p>
</div>
<div class="card card-navy">
<h4>Spending & History</h4>
<p>Income and TotalSpent rank just below Recency, strongly supported by past campaign acceptance history.</p>
</div>
<div class="card card-purple">
<h4>Demographics Lag Behind</h4>
<p>Age sits at the bottom. The model explicitly confirms that behavior predicts conversion better than profile.</p>
</div>
</div>
</div>

## Model Comparison: 5-Fold CV AUC-ROC

<div class="grid-60-40">
<div>
```{r model-compare, fig.height=5.2, fig.width=5.5}
results <- resamples(list(Logistic=log_model, Lasso=lasso_model,
                          RandomForest=rf_model, ANN=ann_model))

roc_df <- results$values %>%
  select(-Resample) %>%
  pivot_longer(cols=everything(), names_to="key", values_to="value") %>%
  separate(key, into=c("Model","Metric"), sep="~") %>%
  filter(Metric=="ROC")

ggplot(roc_df, aes(x=reorder(Model, value, median), y=value, fill=Model)) +
  geom_boxplot(width=0.5, alpha=0.85, outlier.shape=21, outlier.size=2) +
  scale_fill_manual(values=c(Logistic="#94A3B8", Lasso="#64748B",
                              ANN="#7C3AED", RandomForest="#0D9488")) +
  coord_flip() +
  scale_y_continuous(limits=c(0.6,1)) +
  labs(title=NULL, x=NULL, y="AUC-ROC Score") +
  theme_minimal(base_size=14) +
  theme(legend.position="none", plot.background = element_blank(), panel.grid.major.y=element_blank())
```
</div>
<div>
<div class="card card-teal">
<h4>Random Forest Wins</h4>
<p>Highest median AUC and the tightest interquartile range across folds, proving to be the most accurate and consistent model.</p>
</div>
<div class="card card-purple">
<h4>ANN is Unstable</h4>
<p>Similar median AUC but shows much higher fold-to-fold variance, making it less reliable in production.</p>
</div>
<div class="card card-navy">
<h4>Linear Models Fail</h4>
<p>Logistic and Lasso trail consistently. The performance gap confirms the decision boundary is non-linear.</p>
</div>
</div>
</div>

## Test Set Evaluation: Threshold 0.5 vs 0.3

<div class="col2">
<div>
<div class="card card-amber">
<h4>Default Threshold (0.5)</h4>
<p>Highly specific but far too conservative. We accurately identify non-buyers but capture only 18% of actual converters.</p>
</div>
<div class="cm-wrap">
<table class="cm-table">
<tr><th></th><th>Predicted No</th><th>Predicted Yes</th></tr>
<tr><td><b>Actual No</b></td><td class="cm-tn">`r conf_mat$table[1,1]`</td><td class="cm-fp">`r conf_mat$table[2,1]`</td></tr>
<tr><td><b>Actual Yes</b></td><td class="cm-fn">`r conf_mat$table[1,2]`</td><td class="cm-tp">`r conf_mat$table[2,2]`</td></tr>
</table>
</div>
<div style="margin-top:8px; text-align:center;">
<span class="pill pill-red">Sensitivity: `r round(conf_mat$byClass["Sensitivity"], 2)`</span>
<span class="pill pill-teal">Specificity: `r round(conf_mat$byClass["Specificity"], 2)`</span>
</div>
</div>
<div>
<div class="card card-teal">
<h4>Optimized Threshold (0.3)</h4>
<p>Sensitivity skyrockets. We recover the vast majority of potential buyers, accepting a slight drop in specificity.</p>
</div>
<div class="cm-wrap">
<table class="cm-table">
<tr><th></th><th>Predicted No</th><th>Predicted Yes</th></tr>
<tr><td><b>Actual No</b></td><td class="cm-tn">`r conf_adj$table[1,1]`</td><td class="cm-fp">`r conf_adj$table[2,1]`</td></tr>
<tr><td><b>Actual Yes</b></td><td class="cm-fn">`r conf_adj$table[1,2]`</td><td class="cm-tp">`r conf_adj$table[2,2]`</td></tr>
</table>
</div>
<div style="margin-top:8px; text-align:center;">
<span class="pill pill-red">Sensitivity: `r round(conf_adj$byClass["Sensitivity"], 2)`</span>
<span class="pill pill-teal">Specificity: `r round(conf_adj$byClass["Specificity"], 2)`</span>
</div>
</div>
</div>

<div class="hi" style="margin-top:16px;">
<b>Business Logic:</b> Sending a digital offer costs essentially zero. Accepting more false positives to catch real converters maximizes ROI.
</div>

## End-to-end pipeline

<div class="pipe">
<div class="pipe-box" style="border-top:3px solid #1C2B4A">
<h4>1. Segment</h4>
<p>PCA + K-Means splits into 3 economic tiers.</p>
</div>
<div class="pipe-arrow">&rarr;</div>
<div class="pipe-box" style="border-top:3px solid #0D9488">
<h4>2. Score</h4>
<p>Random Forest evaluates conversion probability.</p>
</div>
<div class="pipe-arrow">&rarr;</div>
<div class="pipe-box" style="border-top:3px solid #7C3AED">
<h4>3. Tune</h4>
<p>Lower the decision threshold to 0.3.</p>
</div>
<div class="pipe-arrow">&rarr;</div>
<div class="pipe-box" style="border-top:3px solid #B45309">
<h4>4. Target</h4>
<p>Send customized offers to high-probability converters.</p>
</div>
</div>

<div class="hi" style="margin-top:25px; font-size:14px; text-align:center;">
<b>Result:</b> Marketing budget concentrates strictly on customers who are both in a high-value segment and predicted to convert.
</div>

## Key takeaways

<div class="card card-teal" style="margin-bottom:15px">
<h4>01: Ditch the blanket approach</h4>
<p>Clustering reveals three economically distinct tiers. Segment-specific strategies are the logical baseline, targeting Premium buyers and Browsers with different incentives.</p>
</div>

<div class="card card-navy" style="margin-bottom:15px">
<h4>02: Behavior predicts conversion better than demographics</h4>
<p>Recency and past campaign acceptance are far stronger predictors than age or marital status. Marketing efforts must prioritize recent engagement signals.</p>
</div>

<div class="card card-purple">
<h4>03: Threshold tuning guarantees ROI</h4>
<p>Deployed at an adjusted threshold of 0.3, the Random Forest model becomes a cost-effective screening tool that recovers a large share of potential revenue with minimal overhead.</p>
</div>